---
title: "Creating daily time series using GAM (input data of NPZ model)"
author: "Steven Pint, Viviana Otero, Patricia Cabrera and Salvador Fernandez"
date: "10/12/2021"
output: html_document
---

## Introduction
This document provides a script to create daily time series using Generalized Additive Models (GAM) and the Rpackage 'lwdataexplorer'.

## GAM models
First, we set the directory and load the packages and functions that we are going to use.
```{r }
# Load packages for creating GAM models
source("~/workspace/NPZ/Rscripts/GAM/function_installed_packages_GAM.R")
install_needed_packages_GAM()

# Load functions for creating GAM models
source("~/workspace/NPZ/Rscripts/GAM/functions_GAM_v1_SP.R")
source("~/workspace/NPZ/Rscripts/R scripts/GAM/LW_DataAccess_functions_for_GAM.R", encoding = "UTF-8")
```

Then we provide the needed information to extract data from the LifeWatch database. This information is also used by the GAMs.
```{r }
# Load LifeWatch (LW) data with LW package
start_date = "2011-01-01"
stop_date = Sys.Date()
station_selection = "330"   # c("130","330","LW01", "LW02", "W10", "W09","w09", "W08", "421", "435")
station_region = "midshore"  # c("nearshore","midshore","offshore","offshore","offshore","offshore","offshore","offshore","offshore","offshore")
days_run = as.numeric(difftime(as.Date(stop_date) , as.Date(start_date))) + 1 # how many days are in the run = diff (start - stop) + 1
```

Extract the data needed for the GAMs from LifeWatch database.
```{r }            
station_data <- get_abiotic_pigment(start_date, stop_date, station_selection, station_region)
```


The best GAM model, meaning the model with the lowest Akaike Information Criteria (AIC), is selected to predict and create daily time series for the abiotic factors, i.e. nutrients and SST. Simple visualization is provided for a quick check of the model. 
```{r }  
#This line is for visualization purposes of the performance of the models
par(mfrow=c(2,2))

#GAM PO4
#The selected model has to be defined as mod_p_select
####################################################################

po4 <- station_data[!is.na(station_data$PO4),]
day_station<-yday(po4$Date)
year_station<-year(po4$Date)

models <- list()
models[[1]] <- test_model(po4$PO4, day_station, year_station, 3, 3)
models[[2]] <- test_model(po4$PO4, day_station, year_station, 4, 4)
models[[3]] <- test_model(po4$PO4, day_station, year_station, 6, 6)
AICs <- c(models[[1]]$aic,
          models[[2]]$aic,
          models[[3]]$aic)

mod_p_select <- models[[which.min(AICs)]]

#Comparison observations vs. predictions (by points)
#PO4
plot_po4_1 <- comparisonPoint(day_station, year_station, po4$PO4, mod_p_select)
plot_po4_1 + ylim(0,max(po4$PO4)) + xlim(0,max(po4$PO4)) + labs(x = "PO4 observed", y = "PO4 predicted") 
```
```{r }  
#GAM NH4
#The selected model has to be defined as mod_nh4_select
####################################################################
nh4 <- station_data[!is.na(station_data$NH4),]
day_station<-yday(nh4$Date)
year_station<-year(nh4$Date)

models <- list()
models[[1]] <- test_model(nh4$NH4, day_station, year_station, 3, 3) 
models[[2]] <- test_model(nh4$NH4, day_station, year_station, 4, 4)
models[[3]] <- test_model(nh4$NH4, day_station, year_station, 5, 5)
AICs <- c(models[[1]]$aic,
          models[[2]]$aic,
          models[[3]]$aic)

mod_nh4_select <- models[[which.min(AICs)]]

#Comparison observations vs. predictions
plot_nh4_1 <- comparisonPoint(day_station, year_station, nh4$NH4, mod_nh4_select)
plot_nh4_1 + ylim(0,6) + xlim(0,6) + labs(x = "NH4 observed", y = "NH4 predicted") 
```
```{r }
#GAM NO2
#The selected model has to be defined as mod_no2_select
####################################################################
no2 <- station_data[!is.na(station_data$NO2),]
day_station<-yday(no2$Date)
year_station<-year(no2$Date)

models <- list()
models[[1]] <- test_model(no2$NO2, day_station, year_station, 3, 3)
models[[2]] <- test_model(no2$NO2, day_station, year_station, 4, 4) 
models[[3]] <- test_model(no2$NO2, day_station, year_station, 6, 6)
AICs <- c(models[[1]]$aic,
          models[[2]]$aic,
          models[[3]]$aic)

mod_no2_select <- models[[which.min(AICs)]]

#Comparison observations vs. predictions
plot_no2_1 <- comparisonPoint(day_station, year_station, no2$NO2, mod_no2_select)
plot_no2_1 + ylim(0,1) + xlim(0,1) + labs(x = "NO2 observed", y = "NO2 predicted") 
```
```{r }
#GAM NO3
#The selected model has to be defined as mod_no3_select
####################################################################
no3 <- station_data[!is.na(station_data$NO3),]
day_station<-yday(no3$Date)
year_station<-year(no3$Date)

models <- list()
models[[1]] <- test_model(no3$NO3, day_station, year_station, 3, 3)
models[[2]] <- test_model(no3$NO3, day_station, year_station, 4, 4)
models[[3]] <- test_model(no3$NO3, day_station, year_station, 6, 6)
AICs <- c(models[[1]]$aic,
          models[[2]]$aic,
          models[[3]]$aic)

mod_no3_select <- models[[which.min(AICs)]]

#Comparison observations vs. predictions
plot_no3_1 <- comparisonPoint(day_station, year_station, no3$NO3, mod_no3_select)
plot_no3_1 + ylim(0,20) + xlim(0,20) + labs(x = "NO3 observed", y = "NO3 predicted") 
```
```{r }
#GAM for Silica
#The selected model has to be defined as mod_si_select
#####################################################################################
sil <- station_data[!is.na(station_data$SiO4),]
day_station<-yday(sil$Date)
year_station<-year(sil$Date)

models <- list()
models[[1]] <- test_model(sil$SiO4, day_station, year_station, 3, 3)
models[[2]] <- test_model(sil$SiO4, day_station, year_station, 4, 4)
models[[3]] <- test_model(sil$SiO4, day_station, year_station, 5, 5)
AICs <- c(models[[1]]$aic,
          models[[2]]$aic,
          models[[3]]$aic)

mod_si_select <- models[[which.min(AICs)]]

#Comparison per point
#SiO4
plot_si_1 <- comparisonPoint(day_station, year_station, sil$SiO4, mod_si_select)
plot_si_1 + ylim(0,6) + xlim(0,6) + labs(x = "SiO4 observed", y = "SiO4 predicted") 
```
```{r }
#GAM for Temperature
#The selected model has to be defined as mod_temp_select
#####################################################################################
if (station_region[1] != "offshore"){
temperature <- station_data[!is.na(station_data$Temperature),]
day_station<-yday(temperature$Date)
year_station<-year(temperature$Date)

models <- list()
models[[1]] <- test_model(temperature$Temperature, day_station, year_station, 3, 3)
models[[2]] <- test_model(temperature$Temperature, day_station, year_station, 4, 4) 
models[[3]] <- test_model(temperature$Temperature, day_station, year_station, 5, 5)
models[[4]] <- test_model(temperature$Temperature, day_station, year_station, 6, 6) 
models[[5]] <- test_model(temperature$Temperature, day_station, year_station, 7, 7)
AICs <- c(models[[1]]$aic,
          models[[2]]$aic,
          models[[3]]$aic,
          models[[4]]$aic,
          models[[5]]$aic)

mod_temp_select <- models[[which.min(AICs)]]

#Comparison per point
#temperature
plot_temp <- comparisonPoint(day_station, year_station, temperature$Temperature, mod_temp_select)
plot_temp + ylim(0,25) + xlim(0,25) + labs(x = "Temperature observed", y = "Temperature predicted") 
}
```

## Create daily time series
Now that the best models are selected, it's time to create daily time series for our input data.
```{r }
## Create prediction for NPZD model
##dummy years from 2011 to 2013 will be a copy of the data from 2014 or 2012 depending on availability of data
######################################################################################
predictions <- data.frame(matrix(nrow = days_run, ncol = 7))

names(predictions)[1] <- "day_130"
names(predictions)[2] <- "year_130"
names(predictions)[3] <- "DIN"
names(predictions)[4] <- "Temp"
names(predictions)[5] <- "po4"
names(predictions)[6] <- "sio4"
names(predictions)[7] <- "Date"
```

In the next section, we will use a function to download SST from Meetnet Vlaamse Banken (MVB). To do this, you need to be registered through this link and provide your username and password when asked. This daily SST is used to create daily time series for the nutrient and SST for nearshore and midshore region. In case of offshore region, we use the SST of MVB itself.
```{r }
#Temperature data from MVB
temp_data  <- get_MVB_SST(start_date,stop_date)
colnames(temp_data) <- c("Date", "temperature")
```

```{r }
#Copy dates from MVB data
predictions$Date <- temp_data$Date
predictions$day_130 <- yday(predictions$Date)
predictions$year_130 <- year(predictions$Date)

if (station_region[1] == "offshore"){

  #Add temperature data from MVB directly, here there are not GAM applied
  predictions$Temp <- temp_data$temperature

  #Add values from GAM models
  ##Dummy years are a copy of 2014, new reference dates data frame is created to make the predictions
  #DIN data are only available form 2014 onward
  temp_predictions <- predictions[,c(1,2,7)]
  temp_predictions$year_130[c(1:1096)] <- rep(2014,times =1096)

  #DIN
  predictions[3]<-predict.gam(mod_nh4_select, newdata = temp_predictions[,c(1,2)], type = "response") +
  predict.gam(mod_no2_select, newdata = temp_predictions[,c(1,2)], type = "response") +
  predict.gam(mod_no3_select, newdata = temp_predictions[,c(1,2)], type = "response")

  #PO4
  #Observations of PO4 and SiO4 started in 2012 for offshore stations
  #Therefore, the predictions repeat the year 2012 to fill in the year 2011
  temp_predictions <- predictions[,c(1,2,7)]
  temp_predictions$year_130[c(1:365)] <- rep(2012,times = 365)
  predictions[5]<-predict.gam(mod_p_select, newdata = temp_predictions[,c(1,2)], type = "response") 

  #SiO4
  predictions[6]<-predict.gam(mod_si_select, newdata = temp_predictions[,c(1,2)], type = "response") 

}else{
  
  #Add values from GAM models
  
  #DIN
  #There is no data available from 2011 to 2013 for DIN (see station data)
  #Therefore, 2014 is used as reference year for the dummy years
  #It is repeated 1096 times = 3 years * 365 + 1 day of leap year (2012)
  
  #Create a temporal data frame to calculate the predictions based on the best GAM 
  #select day, year and exact date
  temp_predictions <- predictions[,c(1,2,7)]
  temp_predictions$year_130[c(1:1096)] <- rep(2014,times =1096)

  #Add NH4, NO3, NO2 values based on the exact dates
  predictions[3]<-predict.gam(mod_nh4_select, newdata = temp_predictions[,c(1,2)], type = "response") +
  predict.gam(mod_no2_select, newdata = temp_predictions[,c(1,2)], type = "response") +
  predict.gam(mod_no3_select, newdata = temp_predictions[,c(1,2)], type = "response")


  #Temperature
  #There is temperature data on 2011, therefore it is not necessary to do make a temporal predictions data frame
  predictions[4]<-predict.gam(mod_temp_select, newdata = predictions[,c(1,2)], type = "response") 

  #PO4

  predictions[5]<-predict.gam(mod_p_select, newdata = predictions[,c(1,2)], type = "response") 

  #SiO4

  predictions[6]<-predict.gam(mod_si_select, newdata = predictions[,c(1,2)], type = "response") 
}
```
Let's look how well our predictions are against observations.
```{r }
##Monthly data visualization
########################################################
predictions$month <- month(predictions$Date)
station_data$month <- month(station_data$Date)

comparison_graphs <- compareGraphs(station_data, predictions, days_run)
#DIN
comparison_graphs[[1]]

#PO4
comparison_graphs[[2]]

#SiO4
comparison_graphs[[3]]
```
Let's look at our time series.
```{r }
#compare predictions as time series
comparison_time <- comparisonTimeSeries(predictions, station_data)
#DIN
comparison_time[[1]]

#PO4
comparison_time[[2]]

#SiO4
comparison_time[[3]]
```



```{r }
# adjust negative DIN values to small concentration.
if (length(which(predictions$DIN < 0 )) != 0) {
neg_values_DIN <- which(predictions$DIN < 0 )

min_values <- station_data$DIN[order(station_data$DIN, decreasing=F)]
min_values <- na.omit(min_values)[1:20]
min_DIN <- min(min_values)
max_DIN <- max(min_values)

predictions$DIN[neg_values_DIN] <- runif(length(neg_values_DIN), min_DIN, max_DIN)
}

# adjust negative values to small concentration.
if (length(which(predictions$po4 < 0 )) != 0) {
  neg_values_po4 <- which(predictions$po4 < 0 )
  
  min_values <- station_data$PO4[order(station_data$PO4, decreasing=F)]
  min_values <- na.omit(min_values)[1:20]
  min_po4 <- min(min_values)
  max_po4 <- max(min_values)
  
  predictions$po4[neg_values_po4] <- runif(length(neg_values_po4), min_po4, max_po4)
}

# adjust negative values to small concentration.
if (length(which(predictions$sio4 < 0 )) != 0) {
  neg_values_sio4 <- which(predictions$sio4 < 0 )
  
  min_values <- station_data$SiO4[order(station_data$SiO4, decreasing=F)]
  min_values <- na.omit(min_values)[1:20]
  min_sio4 <- min(min_values)
  max_sio4 <- max(min_values)
  
  predictions$sio4[neg_values_sio4] <- runif(length(neg_values_sio4), min_sio4, max_sio4)
}
```

Now, we write a csv file with our daily time series to the input data folder. This csv file will be used to as input for the NPZ model.
```{r }
#Save predictions as csv
write.csv(predictions, file = paste0("~/workspace/NPZ/Input data/NPZ/inputData_npzd_station", station_region[1],".csv"),row.names = F)
```

